Covariate analysis of CMC RNASeq data reprocessed with RSEM-feature counts workflow with GRCh38 (DLPFC with CQN normlaisation)

### 2018-10-23 code/Rmd/CMC_DLPFC_RSEM_GRCh38_CQN_ModSel.Rmd
Fix metadata

@@

library(ggplot2)
library(limma)
library(psych)
library(edgeR)
library(cqn)
library(lme4)
library(biomaRt)

@@

# Github link
thisRepo <- getRepo(repository = "kelshmo/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain effective count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
downloadFile <- function(id){
  fread(synGet(id)$path, data.table = F)
}

# Download reprocessed counts (DLPFC)
COUNT_ID = 'syn16783526'
ALL_USED_IDs = COUNT_ID
COUNT = downloadFile(COUNT_ID) %>% data.frame()
rownames(COUNT) = COUNT$gene_id
COUNT$gene_id = NULL
COUNT$transcript_id.s. = NULL

@@

# Get RNASeq alignment qc metrics from synapse
METADATA_ALGMETRIC_DLPFC_ID = 'syn16783867'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ALGMETRIC_DLPFC_ID
METADATA_ALGMETRIC_DLPFC = downloadFile(METADATA_ALGMETRIC_DLPFC_ID) %>%
  dplyr::rename(Sample_RNA_ID = ID)

# Get RNASeq QCmetadata
METADATA_QC_DLPFC_ID = 'syn16816488'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_DLPFC_ID
METADATA_QC_DLPFC = downloadFile(METADATA_QC_DLPFC_ID) %>%
  dplyr::select(Individual_ID, Institution, Gender, Ethnicity, Age_of_Death, `PMI_(in_hours)`, Dx,
                Sample_RNA_ID, one_of('rnaSeq_isolation:Exclude?', 'rnaSeq_isolation:RIN',
                                      'rnaSeq_report:Ribozero_Batch', 'rnaSeq_report:Library_Batch', 
                                      'rnaSeq_report:Flowcell_Batch', 'rnaSeq_report:Exclude?'))
```

### Data preprocessing
```{r preprocess.data}
# Merge metadata
METADATA = METADATA_QC_DLPFC %>%
  dplyr::left_join(GENOTYPE) %>%
  dplyr::left_join(METADATA_ALGMETRIC_DLPFC) %>%
  dplyr::rename(PMI = `PMI_(in_hours)`,
                IsolationExclude = `rnaSeq_isolation:Exclude?`, 
                RIN = `rnaSeq_isolation:RIN`, 
                ReportExclude = `rnaSeq_report:Exclude?`,
                SampleID = Sample_RNA_ID,
                LibraryBatch = `rnaSeq_report:Library_Batch`,
                FlowcellBatch = `rnaSeq_report:Flowcell_Batch`,
                RibozeroBatch = `rnaSeq_report:Ribozero_Batch`,
                MappedReads = Mapped,
                IntragenicRate = `Intragenic Rate`, IntronicRate = `Intronic Rate`, 
                IntergenicRate = `Intergenic Rate`, GenesDetected = `Genes Detected`,
                ExpProfEfficiency = `Expression Profiling Efficiency`, rRNARate = `rRNA rate`) %>%
  dplyr::select(SampleID, Individual_ID, Institution, Gender, Ethnicity, Age_of_Death, PMI, Dx, 
                RIN, IsolationExclude, ReportExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
                LibraryBatch, FlowcellBatch, RibozeroBatch, MappedReads, TotalReads, GenesDetected, AlignmentRate,
                IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate) %>%
  dplyr::filter(SampleID %in% colnames(COUNT), !is.na(SampleID)) %>%
  dplyr::mutate(Dx = forcats::fct_recode(Dx,  Other = "AFF", Other = "BP", Control = "Control", SCZ = "SCZ"))

@@

ind = METADATA$SampleID [is.na(METADATA$Gender)]
writeLines(paste('Following', length(ind), 'counts are missing gender'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

@@

GENE.GC.CONT = Ensemble2HGNC %>%
  dplyr::left_join(backgroundGenes) %>%
  dplyr::select(gene_id, percentage_gene_gc_content) %>%
  unique
rownames(GENE.GC.CONT) = GENE.GC.CONT$gene_id
```

### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c('Individual_ID', "Institution", "Gender", "LibraryBatch", "Dx", "FlowcellBatch")
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", 
                    "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

@@

# Gender
# vcd::mosaic(~ Gender + Dx, data = COVARIATES)
```
### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx and per Dx.sex. Also remove genes with missing gene length and percentage GC content

@@

# Iteratively identify covariates in phase I
phase1.covars = c('Institution', 'Gender', 'PMI', 'Age_of_Death', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

>>>>

Notes: 
- 3 Kleinfelter individuals

@@

library(ggplot2)
library(limma)
library(psych)
library(edgeR)
library(cqn)
library(lme4)
library(biomaRt)
library(ComplexHeatmap)

@@

# Github link
thisRepo <- getRepo(repository = "kelshmo/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain effective count matrix and metadata from synapse.

#readRDS(file = "logcpm.rds")
```{r download.data, cache=TRUE}
downloadFile <- function(id){
  fread(synGet(id)$path, data.table = F)
}
downloadFile_version <- function(id , version){
  fread(synGet(id, version = version)$path, data.table = F)
}
# Download reprocessed counts (DLPFC)
COUNT_ID = 'syn16783526'
ALL_USED_IDs = COUNT_ID
COUNT = downloadFile(COUNT_ID) %>% data.frame()
rownames(COUNT) = COUNT$gene_id
COUNT$gene_id = NULL
COUNT$transcript_id.s. = NULL

@@

# Get RNASeq QCmetadata
METADATA_QC_DLPFC_ID = 'syn16816488' 
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_DLPFC_ID
METADATA_QC_DLPFC = downloadFile_version(METADATA_QC_DLPFC_ID, version = 7) %>%
  dplyr::select(Individual_ID, Institution, Reported_Gender, Sex, Ethnicity, Age_of_Death, `PMI_(in_hours)`, Dx,
                Sample_RNA_ID, one_of('rnaSeq_isolation:Exclude?', 'rnaSeq_isolation:RIN',
                                      'rnaSeq_report:Ribozero_Batch', 'rnaSeq_report:Library_Batch', 
                                      'rnaSeq_report:Flowcell_Batch', 'rnaSeq_report:Exclude?', 'rnaSeq_report:Mapped_Reads',
                                      'rnaSeq_report:Intragenic_Rate','rnaSeq_report:Intronic_Rate',
                                      'rnaSeq_report:Intergenic_Rate','rnaSeq_report:Genes_Detected',
                                      'rnaSeq_report:Expression_Profiling_Efficiency', 'rnaSeq_report:rRNA_Rate',
                                      'rnaSeq_report:Total_Reads', 'rnaSeq_report:Alignment_Rate'))
```

### Data preprocessing
```{r preprocess.data}
# Merge metadata
METADATA = METADATA_QC_DLPFC %>%
  dplyr::left_join(GENOTYPE) %>%
  dplyr::rename(PMI = `PMI_(in_hours)`,
                IsolationExclude = `rnaSeq_isolation:Exclude?`, 
                RIN = `rnaSeq_isolation:RIN`, 
                ReportExclude = `rnaSeq_report:Exclude?`,
                SampleID = Sample_RNA_ID,
                LibraryBatch = `rnaSeq_report:Library_Batch`,
                FlowcellBatch = `rnaSeq_report:Flowcell_Batch`,
                RibozeroBatch = `rnaSeq_report:Ribozero_Batch`,
                MappedReads = `rnaSeq_report:Mapped_Reads`,
                IntragenicRate = `rnaSeq_report:Intragenic_Rate`, IntronicRate = `rnaSeq_report:Intronic_Rate`, 
                IntergenicRate = `rnaSeq_report:Intergenic_Rate`, GenesDetected = `rnaSeq_report:Genes_Detected`,
                ExpProfEfficiency = `rnaSeq_report:Expression_Profiling_Efficiency`, rRNARate = `rnaSeq_report:rRNA_Rate`,
                TotalReads = `rnaSeq_report:Total_Reads`, AlignmentRate = `rnaSeq_report:Alignment_Rate`) %>%
  dplyr::select(SampleID, Individual_ID, Institution, Reported_Gender,Sex, Ethnicity, Age_of_Death, PMI, Dx, 
                RIN, IsolationExclude, ReportExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
                LibraryBatch, FlowcellBatch, RibozeroBatch, MappedReads, TotalReads, GenesDetected, AlignmentRate,
                IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate) %>%
  dplyr::filter(SampleID %in% colnames(COUNT), !is.na(SampleID)) %>%
  dplyr::mutate(Dx = forcats::fct_recode(Dx,  Other = "AFF", Other = "BP", Control = "Control", SCZ = "SCZ"))

@@

ind = METADATA$SampleID [is.na(METADATA$Reported_Gender)]
writeLines(paste('Following', length(ind), 'counts are missing gender'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

@@

GENE.GC.CONT = Ensemble2HGNC %>%
  dplyr::left_join(backgroundGenes) %>%
  dplyr::select(gene_id, percentage_gene_gc_content, chromosome_name) %>%
  unique
rownames(GENE.GC.CONT) = GENE.GC.CONT$gene_id
```

### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c('Individual_ID', "Institution", "Reported_Gender", "LibraryBatch", "Dx", "FlowcellBatch")
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", 
                    "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

@@

# Gender
# vcd::mosaic(~ Reported_Gender + Dx, data = COVARIATES)
```
### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx and per Dx.sex. Also remove genes with missing gene length and percentage GC content

@@

# Iteratively identify covariates in phase I
phase1.covars = c('Institution', 'Reported_Gender', 'PMI', 'Age_of_Death', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)
